---
title: "How many fishes are in the lake?"
output:
  html_document:
    df_print: paged
---

The problem: how we estimate the size of a fish population in a lake? Obviously without catch all fishes or drain the entire like to find out. Mark and recapture is a method commonly used in ecology to estimate an animal population's size. In this post we simulate the "mark and recatch" technique to assess a population sizes through sampling.

<!--more--> 

**Mark and recapture** is a method commonly used in ecology to estimate an animal population's size. A portion of the population is captured, marked, and released. Later, another portion is captured and the number of marked individuals within the sample is counted. Since the number of marked individuals within the second sample should be proportional to the number of marked individuals in the whole population, an estimate of the total population size can be obtained by dividing the number of marked individuals by the proportion of marked individuals in the second sample. The method is most useful when it is not practical to count all the individuals in the population. [^1]

## Simulation

We'll simulate a cenario with a previously population known where we can use this technique to assess the quality of the information obtained. So, let's create a population and a "lake" (the "fishes vector") to be sampled.

```{r setupLake}

# libraries
library(dplyr)

# generating a fish population between 10k and 50k 
set.seed(1975)
population <- round(runif(1, min=10000, max=50000)) # 'unknown population'
fishes     <- vector(length = population)

```

Now we do the first catch and mark, we sample the lake population and mark the fishes.

```{r mark}

# sample and mark
marked_fishes <- 1000
fishes[sample(1:population, marked_fishes, replace=F)] <- T

```

Now we resample the lake and check how many marked fishes we'll get.

```{r recatch}

# recatch the fishs
sample_size <- 1000
recatch <- fishes[sample(1:population, sample_size, replace=F)]

# check how many fishes are marked fishes
marked_sample <- sum(recatch)
marked_sample
  
```

We recatch `r marked_sample` maked fishs in a sample of `r sample_size`, so we now know that the proportion of marked fishes in the tank are `r round(marked_sample/sample_size,3)`. Knowing that are 1000 marked fishes represents `r round(100*marked_sample/sample_size,1)`% of a total population we can now to estimate the population size.

```{r popEst}

# population estimative
marked_proportion <- marked_sample/sample_size
population_est <- marked_fishes/marked_proportion
round(population_est)

```
And compare with the real value:

```{r compare}

# what is the real population
population

# checking for the error 
abs(population_est-population)/population

```

The number are got was `r as.integer(round(population_est))` against a real value of `r as.integer(population)`, we error the population by `r round(100*abs(population_est-population)/population,1)`%. 

## Simulations and Experiments

Before going to the math (this is basically a well know binary distribution case) let's go some experiment. The size of the samples (the "mark catch" and the "recatch and count") alters the precision of the information? Do several fewer catchs is better than a large catch?

### More fewer catchs is better than a large catch?

Let's see if an strategy of recatch the same number of fishes (`r sample_size`) but in several fewer catchs is better than a unique recatch.

```{r recatchSize}

# sizes recatch samples of the experiments
recatch_sizes <- c(100,200,250,500,1000)

# for each recatch size we make the sampling and mean
results <- lapply(recatch_sizes, function(rs){
  
  # setup the lake and marked fishes
  fishes     <- vector(length = population)
  fishes[sample(1:population, marked_fishes, replace=F)] <- T
  
  # how many times we do a recatch
  repetitions <- sample_size/rs
  sampling_result <- vector(mode = "integer", length = repetitions)
  
  for(i in 1:10){
    
    # recatch the fishs
    recatch <- fishes[sample(1:population, rs, replace=F)]
    # check how many fishes are marked fishes
    marked_sample <- sum(recatch)
    # estimate the population
    marked_proportion  <- marked_sample/sample_size
    sampling_result[i] <- round(marked_fishes/marked_proportion)

  }
  
  # return the mean value
  return(data.frame(
    repetitions = repetitions,
    sample_size = rs,
    pop_est_mean = mean(sampling_result),
    pop_est_sd   = sd(sampling_result)
  ))
  
})

bind_rows(results)
print(population)

```


## References

[^1]: [Wikipedia - Mark and Capture](https://en.wikipedia.org/wiki/Mark_and_recapture)
